\vspace{-1.5em} \bf A general and efficient representation of ancestral recombination graphs

### 2023-10-23 paper.tex
Yans updates

\linenumbers
\section{Introduction}
\label{sec-intro}
% Para 1: ARGs are cool!
Estimating the genetic genealogy of a set of genome sequences
under the influence of recombination,
usually known as an Ancestral Recombination Graph (ARG), is a long-standing
goal in genetics.
Broadly speaking, an ARG describes the different paths of genetic inheritance
caused by recombination, and encodes a sequence of correlated genealogical
trees along the genome.
Recent breakthroughs
in large-scale inference
methods~\citep{rasmussen2014genome,kelleher2019inferring,speidel2019method,
schaefer2021ancestral,wohns2022unified,zhang2023biobank,zhan2023towards}
% and data representation~\citep{kelleher2016efficient,kelleher2018efficient}
have raised the realistic prospect of ARG-based analysis becoming a standard part
of the population and statistical genetics toolkit~\citep{hejase2020summary}.
Applications using inferred ARGs as input have begun to
appear~\citep{osmond2021estimating,
fan2022genealogical,
hejase2022deep,
guo2022recombination,
zhang2023biobank,
nowbandegani2023extremely,
ignatieva2023distribution,
fan2023likelihood}
and many more are sure to
follow~\citep{harris2019database,harris2023using}.
See \citet{lewanski2023era} for a biologically oriented introduction to ARGs
and further discussion on the potential impacts.

@@

% Para 3: what is the problem?
% FIXME current version isn't great, but it's better than previous effort.
[FIXME this paragrph needs a substantial update, need to reread the
eARG and gARG sections again to be clear what we say we're doing.
We should focus on the concrete interchange aspects rather than
coalescenty type things.]
The mismatch between the classical Griffiths definition of an ARG and the
modern interpretation is the source of substantial confusion,
and, we argue, a significant limiting factor for the field.
% actual outputs of these modern methods is a central issue.
As this definition is essentially a reformulation of the
coalescent with recombination (henceforth ``the coalescent''; we
will assume the incorporation of recombination for the purposes of
this article) in graph-theoretic form,
the distinction between the simplifying assumptions of the model
and the necessary requirements for a data structure are not clear.
The consequences of conflating model with data structure
are subtle, varied, and deep.
% TODO rephrase, not great currentyl
We argue that the prevailing definition of an ARG
imposes unneccesary limitations, and that with a broader definition
we can reassess what we should be attempting to estimate,
how we measure success, and the expected limitations on the precision
and accuracy of our inferences.

@@

% TODO: Provide a map of the paper's sections and how they tie together in support of gARG as a way forward.
% Here is a rough draft - it would be good to state the most important message (in one sentence or so) from each of the sections to help with frontloading key messages
% Genome ARGs &
% Event ARGs
In the following, we first define the genome ARG encoding
and contrast it with the event ARG approach.
We then discuss the important concept of ancestral material in
Section~\ref{sec-ancestral-material}
and [DO SOME STUFF].
% Converting and eARG to gARG &
% ARGs and local trees &
% Locally unary nodes &
% Levels of simplification
This simplification is closely related to operations needed to convert eARG and gARG,
which opens a rich set of details about ARGs and local trees, local unary nodes, and
levels of simplification.
% Precision of recombination information
We then highlight the important difference between the required level of precision
about recombination information for eARGs compared to gARGs.
% Computational efficiency and data interchange
The way this recombination information is stored in eARGs and gARGs has important
implications for computational efficiency, which is critical if we want to use ARGs
for modern genomic datasets.
% Discussion
We wrap with a discussion highlighting that WHAT-IS-THE-KEY-SINGLE-MESSAGE-NOT-COVERED-ABOVE?
Finally, the literature on ARGs and associated aspects is vast, particularly
due to different definitions and implementations. We cover some important aspects
in appendices, including
a brief history of ancestral graphs (Appendix~\ref{sec-arg-history}),
description of the big and little ARGs (Appendix~\ref{sec-big-and-little-arg}),
survey of ARG inference methods (Appendix~\ref{sec-survey-arg-infer}), and
ARGs at an individual vs cell lineage level (Appendix~\ref{sec-cell-lineages-and-args}).

@@

% Ordered alphabetically by surname
Gregor Gorjanc acknowledges support from the BBSRC ISP grant to The Roslin Institute
(BBS/E/D/30002275, BBS/E/RL/230001A, and BBS/E/RL/230001C).
% Anastasia Ignatieva acknowledges support from TODO
% TODO Jerome: add EPSRC and NIH grants.
Jerome Kelleher acknowledges support from the Robertson Foundation.
Jere Koskela acknowledges support from EPSRC research grant EP/V049208/1.
% Anthony Wilder Wohns acknowledges support from TODO

>>>>

\linenumbers
\section{Introduction}
\label{sec-intro}
% Para 1: ARGs are cool!
Estimating the genetic genealogy of a set of genome sequences
under the influence of recombination,
usually known as an Ancestral Recombination Graph (ARG), is a long-standing
goal in genetics.
Broadly speaking, an ARG describes the different paths of genetic inheritance
caused by recombination, and encodes a sequence of correlated genealogical
trees along the genome.
Recent breakthroughs
in large-scale inference
methods~\citep{rasmussen2014genome,kelleher2019inferring,speidel2019method,
schaefer2021ancestral,wohns2022unified,zhang2023biobank,zhan2023towards}
% and data representation~\citep{kelleher2016efficient,kelleher2018efficient}
have raised the realistic prospect of ARG-based analysis becoming a standard part
of the population and statistical genetics toolkit~\citep{hejase2020summary}.
Applications using inferred ARGs as input have begun to
appear~\citep{osmond2021estimating,
fan2022genealogical,
hejase2022deep,
guo2022recombination,
zhang2023biobank,
nowbandegani2023extremely,
ignatieva2023distribution,
fan2023likelihood}
and many more are sure to
follow~\citep{harris2019database,harris2023using}.
\citet{lewanski2023era} provide a biologically oriented introduction to ARGs
and further discuss their potential impacts.

@@

% Para 3: what is the problem?
% FIXME current version isn't great, but it's better than previous effort.
[FIXME this paragrph needs a substantial update, need to reread the
eARG and gARG sections again to be clear what we say we're doing.
We should focus on the concrete interchange aspects rather than
coalescenty type things.]
The mismatch between the classical Griffiths definition of an ARG and the
modern interpretation is the source of substantial confusion,
and, we argue, a significant limiting factor for the field.
% actual outputs of these modern methods is a central issue.
As this definition is essentially a reformulation of the
coalescent with recombination (henceforth ``the coalescent''; we
will assume the incorporation of recombination for the purposes of
this article) in graph-theoretic form,
the distinction between the simplifying assumptions of the model
and the necessary requirements for a data structure are not clear.
The consequences of conflating model with data structure
are subtle, varied, and deep.
% TODO rephrase, not great currently
We argue that the prevailing definition of an ARG
imposes unnecessary limitations, and that with a broader definition
we can reassess what we should be attempting to estimate,
how we measure success, and the expected limitations on the precision
and accuracy of our inferences.

@@

% TODO: Provide a map of the paper's sections and how they tie together in support of gARG as a way forward.
% Here is a rough draft - it would be good to state the most important message (in one sentence or so) from each of the sections to help with frontloading key messages
% Genome ARGs &
% Event ARGs
In the following, we first define the genome ARG encoding
and contrast it with the event ARG approach.
We then discuss the important concept of ancestral material in
Section~\ref{sec-ancestral-material}
and [DO SOME STUFF].
% Converting and eARG to gARG &
% ARGs and local trees &
% Locally unary nodes &
% Levels of simplification
This simplification is closely related to operations needed to convert eARGs to
various forms of gARGs,
which opens a rich set of details about ARGs and local trees, local unary nodes, and
levels of simplification.
% Precision of recombination information
We then highlight the important difference between the required level of precision
about recombination information for eARGs compared to gARGs.
% Computational efficiency and data interchange
The way this recombination information is stored in eARGs and gARGs has important
implications for computational efficiency, which is critical if we want to use ARGs
for modern genomic datasets.
% Discussion
We wrap with a discussion highlighting that WHAT-IS-THE-KEY-SINGLE-MESSAGE-NOT-COVERED-ABOVE?
Finally, the literature on ARGs and associated aspects is vast, particularly
due to different definitions and implementations. We cover some important aspects
in appendices, including
a brief history of ancestral graphs (Appendix~\ref{sec-arg-history}),
description of the big and little ARGs (Appendix~\ref{sec-big-and-little-arg}),
survey of ARG inference methods (Appendix~\ref{sec-survey-arg-infer}), and
ARGs at an individual vs cell lineage level (Appendix~\ref{sec-cell-lineages-and-args}).

@@

% Ordered alphabetically by surname
Gregor Gorjanc acknowledges support from the BBSRC ISP grant to The Roslin Institute
(BBS/E/D/30002275, BBS/E/RL/230001A, and BBS/E/RL/230001C).
% Anastasia Ignatieva acknowledges support from TODO
% TODO Jerome: add EPSRC and NIH HG011395-0 grants.
Jerome Kelleher acknowledges support from the Robertson Foundation.
Jere Koskela acknowledges support from EPSRC research grant EP/V049208/1.
% Anthony Wilder Wohns acknowledges support from TODO
